A Logical Solution of the Linearity Challenge

### 2025-12-04 paper/Context.lagda.tex
general cleanup

When splitting a context $\Context$ into $\ContextD + \ContextE$,
for each type in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<\_} and \AgdaInductiveConstructor{>\_} to indicate whether it is meant to be placed in $\ContextD$ or $\ContextE$,
respectively.
%
As an example, below is the splitting of the context $[A, B, C, D]$ into $[B] + [A, C, D]$.

@@

It is often the case that the context $\ContextD$ in a splitting
$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. We use this
particular splitting to describe the presence of a channel of type $A$ in
$\ContextC$ and giving a name $\ContextE$ to the residual context. For this
particular case, we introduce a dedicated notation that allows us to write this
case in a more compact and meaningful way, as
$\SimpleSplitting\ContextC{A}\ContextE$.

@@

Note that, when specifying a splitting $\SimpleSplitting\ContextC{A}\ContextE$,
it may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
applications of \AgdaInductiveConstructor{>\_} after the $A$ type has been
reached. For instance, the same sample splitting shown earlier can be written
equivalently and in a more compact way as shown below. More examples will be
given in \Cref{sec:examples}.

@@

It is a simple fact that the merging of two unrestricted contexts
$\Context_1$ and $\Context_2$ results in an unrestricted context.\Luca{Serve?}

\begin{code}
+-un : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
\end{code}
\begin{code}[hide]
+-un • un-[] un-[] = un-[]
+-un (< p) (un-∷ un₁) un₂ = un-∷ (+-un p un₁ un₂)
+-un (> p) un₁ (un-∷ un₂) = un-∷ (+-un p un₁ un₂)
\end{code}

>>>>

When splitting a context $\Context$ into $\ContextD + \ContextE$, for each type
in $\Context$ we use one of the prefix operators \AgdaInductiveConstructor{<}
and \AgdaInductiveConstructor{>} to indicate whether the type is meant to be
placed in $\ContextD$ or $\ContextE$, respectively.
%
As an example, below is the splitting of the context $[A, B, C, D]$ into $[B] +
[A, C, D]$.

@@

It is often the case that the context $\ContextD$ in a splitting
$\Splitting\ContextC\ContextD\ContextE$ is a singleton list $[A]$. We use this
particular splitting to describe the presence of a channel of type $A$ in
$\ContextC$ and giving a name $\ContextE$ to the residual context. For this
particular case, we introduce the dedicated notation
$\SimpleSplitting\ContextC{A}\ContextE$ which we find clearer.

@@

Note that, when specifying a splitting $\SimpleSplitting\ContextC{A}\ContextE$,
it may be convenient to use \AgdaFunction{≫} as a shortcut for a sequence of
applications of \AgdaInductiveConstructor{>} after the $A$ type has been
reached. For instance, \AgdaFunction{splitting-example₁} can be written
equivalently and in a more compact way as shown below. More usages of
\AgdaFunction{≫} will be given in \Cref{sec:examples}.

@@

% It is a simple fact that the merging of two unrestricted contexts
% $\Context_1$ and $\Context_2$ results in an unrestricted context.\Luca{Serve?}

\begin{code}[hide]
+-un : ∀{Γ Γ₁ Γ₂} → Γ ≃ Γ₁ + Γ₂ → Un Γ₁ → Un Γ₂ → Un Γ
+-un • un-[] un-[] = un-[]
+-un (< p) (un-∷ un₁) un₂ = un-∷ (+-un p un₁ un₂)
+-un (> p) un₁ (un-∷ un₂) = un-∷ (+-un p un₁ un₂)
\end{code}

### 2025-12-04 paper/Permutations.lagda.tex
restored description of Process
general cleanup

Just like permutations from Agda's standard library, also \AgdaDatatype{\_↭\_}
enjoys a number of expected properties, such as symmetry. Among them, in the
following we will make use of a specific one that connects context splitting and
permutations: if $\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a
permutation of the concatenation of $\ContextD$ and $\ContextE$.\Luca{Ci sono
molte altre proprietà non descritte, controllare quelle che servono davvero.}

@@

↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ≃ A , Γ′ → ∃[ Δ′ ] (Δ ≃ A , Δ′ × Γ′ ↭ Δ′)
↭solo π p with ↭split π p
... | _ , _ , q , π₁ , π₂ rewrite ↭solo-inv π₁ = _ , q , π₂

>>>>

Just like permutations from Agda's standard library, also \AgdaDatatype{\_↭\_}
enjoys a number of expected properties, such as symmetry. Among them, in the
following we will make use of a specific property that relates context
splitting, list concatenation and permutations: if
$\Splitting\ContextC\ContextD\ContextE$, then $\ContextC$ is a permutation of
the concatenation of $\ContextD$ and $\ContextE$.

@@

↭solo : ∀{A Γ Γ′ Δ} → Γ ↭ Δ → Γ ∋ A ⊳ Γ′ → ∃[ Δ′ ] (Δ ∋ A ⊳ Δ′ × Γ′ ↭ Δ′)
↭solo π p with ↭split π p
... | _ , _ , q , π₁ , π₂ rewrite ↭solo-inv π₁ = _ , q , π₂

### 2025-12-04 paper/Reduction.lagda.tex
general cleanup

The remaining constructors that describe the base reductions follow a similar
pattern. We only glance at \RExists since it involves a non-trivial rewriting of
types:

\begin{code}
  r-ex         : ∀{A B Γ₁ Γ₂ P F}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                 cut {`∃ A} p (ex {_} {B} (< p₀) P) (all (< q₀) F) ↝ cut p P (F B)
\end{code}

>>>>

The remaining constructors that describe the base reductions follow a similar
pattern, noting that the implementation of \RWeaken and \RContract makes use of
auxiliary functions that respectively weaken and contract the typing context of
the resulting process as shown in \Cref{tab:semantics}. It is worth glancing at
\RExists since it involves a non-trivial rewriting of types:

\begin{code}
  r-exists     : ∀{A B Γ₁ Γ₂ P F}
                 (p : Γ ≃ Γ₁ + Γ₂) (p₀ : Γ₁ ≃ [] + Γ₁) (q₀ : Γ₂ ≃ [] + Γ₂) →
                 cut {`∃ A} p (ex {_} {B} (< p₀) P) (all (< q₀) F) ↝ cut p P (F B)
\end{code}
